### 2014-05-04 stat246/hw4.tex
s4 done

The quantity to maximize is now $l(\theta|X,Y)+\log p(\theta)$, which in this case is
\begin{gather*}
    \sum_{i=1}^N\sum_{m=1}^Mw_{i,m}\left(\log\pi_m+\sum_{j=1}^dx_{i,j}\log p_{j,m}+(1-x_{i,j})\log(1-p_{j,m})\right)\\
    +\sum_{i=1}^N\sum_{m=1}^M\sum_{j=1}^d\frac{1}{N}\left((\alpha-1)\log p_{j,m}+(\alpha-1)\log(1-p_{j,m})\right)\\
    +\frac{1}{N}\sum_{i=1}^N\sum_{m=1}^M(\zeta_m-1)\log\pi_m
\end{gather*}
after removing constants irrelevant to maximization. Folding the sums in gives
\[\sum_{i=1}^N\sum_{m=1}^M\left(w_{i,m}+\frac{1}{N}(\zeta_m-1)\right)\log\pi_m+\sum_{i=1}^N\sum_{m=1}^Mw_{i,m}\sum_{j=1}^d\left(x_{i,j}+\frac{1}{N}(\alpha-1)\right)\log p_{j,m}+\left(1-x_{i,j}+\frac{1}{N}(\alpha-1)\right)\log(1-p_{j,m})\]
Having the same form as the thing we maximized already, we find that the new equations for $\pi_m$ and $p_{j,m}$ are
\begin{align*}
    \pi_m&=\kappa\sum_{i=1}^N\left(w_{i,m}+\frac{1}{N}(\zeta_m-1)\right)\\
    p_{j,m}&=\frac{\sum_{i=1}^Nw_{i,m}\left(x_{i,j}+\frac{1}{N}(\alpha-1)\right)}{\left(\frac{2}{N}(\alpha-1)+1\right)\sum_{i=1}^Nw_{i,m}}
\end{align*}
where $\kappa$ is some constant that makes the $\pi_m$ sum to 1.

For computational purposes in computing the $w_{i,m}$, instead of computing
\[\frac{P(x_i|Y=m)P(Y=m)}{\sum_{j=1}^MP(x_i|Y=j)P(Y=j)}\]
compute instead
\[\log (P(x_i|Y=m)P(Y=m))=\sum_{j=1}^d\left(x_{i,j}\log p_{j,m}+(1-x_{i,j})\log(1-p_{j,m})\right)+\log\pi_m\]
for each pair $i,m$ and call them the $\alpha_{i,m}$. For each $i$, calculate $l_i=\max_m\alpha_{i,m}$, and set $w_{i,m}=\frac{\exp(\alpha_{i,m}-l_i)}{\sum_{m'}\exp(\alpha_{i,m'}-l_i)}$.
\end{document}

>>>>

The quantity to maximize is now $l(\theta|X,Y)+\log p(\theta)$, which in this case is
\begin{gather*}
    \sum_{i=1}^N\sum_{m=1}^Mw_{i,m}\left(\log\pi_m+\sum_{j=1}^dx_{i,j}\log p_{j,m}+(1-x_{i,j})\log(1-p_{j,m})\right)\\
    +\sum_{i=1}^N\sum_{m=1}^M\sum_{j=1}^d\frac{1}{N}\left((\alpha-1)\log p_{j,m}+(\alpha-1)\log(1-p_{j,m})\right)\\
    +\frac{1}{N}\sum_{i=1}^N\sum_{m=1}^M(\zeta_m-1)\log\pi_m
\end{gather*}
after removing constants irrelevant to maximization. Folding the sums in gives
\[\sum_{i=1}^N\sum_{m=1}^M\left(w_{i,m}+\frac{1}{N}(\zeta_m-1)\right)\log\pi_m+\sum_{i=1}^N\sum_{m=1}^Mw_{i,m}\sum_{j=1}^d\left(x_{i,j}+\frac{1}{N}(\alpha-1)\right)\log p_{j,m}+\left(1-x_{i,j}+\frac{1}{N}(\alpha-1)\right)\log(1-p_{j,m})\]
Having the same form as the thing we maximized already, we find that the new equations for $\pi_m$ and $p_{j,m}$ are
\begin{align*}
    \pi_m&=\kappa\sum_{i=1}^N\left(w_{i,m}+\frac{1}{N}(\zeta_m-1)\right)\\
    p_{j,m}&=\frac{\sum_{i=1}^Nw_{i,m}\left(x_{i,j}+\frac{1}{N}(\alpha-1)\right)}{\left(\frac{2}{N}(\alpha-1)+1\right)\sum_{i=1}^Nw_{i,m}}
\end{align*}
where $\kappa$ is some constant that makes the $\pi_m$ sum to 1. In practice, all the $\zeta_m$ were set to $2$, as were all the $\alpha$s

For computational purposes in computing the $w_{i,m}$, instead of computing
\[\frac{P(x_i|Y=m)P(Y=m)}{\sum_{j=1}^MP(x_i|Y=j)P(Y=j)}\]
compute instead
\[\log (P(x_i|Y=m)P(Y=m))=\sum_{j=1}^d\left(x_{i,j}\log p_{j,m}+(1-x_{i,j})\log(1-p_{j,m})\right)+\log\pi_m\]
for each pair $i,m$ and call them the $\alpha_{i,m}$. For each $i$, calculate $l_i=\max_m\alpha_{i,m}$, and set $w_{i,m}=\frac{\exp(\alpha_{i,m}-l_i)}{\sum_{m'}\exp(\alpha_{i,m'}-l_i)}$.

Now for the results. The fitted probability vectors are shown below:

\includegraphics[width=0.9\textwidth]{hw4_files/M1_princip.png}

\includegraphics[width=0.9\textwidth]{hw4_files/M3_princip.png}

\includegraphics[width=0.9\textwidth]{hw4_files/M5_princip.png}

In classifying the test set, I went class-by-class and computed the likelihoods of each data vector given the parameters that were fitted to a certain class. The likelihoods were computed by exponentiating the log-likelihood, which is more numerically stable than just multiplying together the 900 Bernoulli likelihoods. The error rates observed were $0.2041$ for $M=1$, $0.1396$ for $M=3$, and $0.1383$ for $M=5$. This indicates that the digits can already be described quite well by a mixture of three components (which, looking at the images above, correspond to variations in how people draw the digits).
\end{document}
